---
title: "Foundations of R programming I"
author: "Shamit Soneji"
date: "12/10/2018"
output: rmarkdown::github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Why use R?

R is a programming environment with a focus on mathematics and statistics, but can be used for a variety of applications given the flexibility of the language. R is also free and available for all operating systems. Given the richness of the language and no cost to use it, bioinformaticians have adopted R as the platform for which which to develop packages to solve bioinformatics problems.

## Getting R

Point your browser to http://cran.r-project.org/ to download and install the latest version of R. For these tutorials we are also going o use [RStudio](http://www.rstudio.com/) which is an advanced environment for R which includes a window for an editor, console, and plotting window. You will see what this means later.

With respect to bioinformatics, the central repository for bioinformatics tools is [Bioconductor](http://www.bioconductor.org) where packages are deposited for easy installation.

Before we go on to these, we need to get to grips with the basics of the R language first.

## R- The basics

### 1- Vectors

The R language is relatively intuative. For example, making a string of numbers from 1 to 10:

```{r}
x <- c(1,2,3,4,5,6,7,8,9,10)
```
The "c" in the code above means "combine", therefore all the comma separated numbers between the parentheses are put together to create a `vector`.

To view the contents of the object you have just created, just type "x" and hit return:

```{r}
x
```

There is a much simpler way to create the same type of object:

```{r}
x <- 1:10
x
```
Much better. Using a comma will always do increments of 1, but is also bidirectional:

```{r}
y <- 5:-5
y
```
Another way of ceating a sequence of numbers is to use the `seq` function. To learn how this function works, issue the command `help(seq)`. In R you can get a manual for any function using the `help()` command. To generate a vector of numbers from 1 to 100 in steps of 10 we need:

```{r}
a <- seq(0,100,by=10)
a
```
***Exercise:*** Generate a vector called 'b' ranging from 3 to 987 where the length of the vector is 53 entries long.
Done? Check the length of the vector you have just made by issuing `length(b)`.

Now that we can make vectors we can start playing with them. for example:
```{r}
c <- 1:50
d <- 1/c
```
Lets plot the numbers contained in the object we called `d`:

```{r fig.width=7, fig.height=6}
plot(d)
```
Note the way the axes are labelled in the plot function.

***Exercise:*** Call `help(plot)` and read about the other options available. Produce the same plot as above, but this time as a line plot which is coloured red. Also, label the axes and give the plot a title.

We can also do basic calculations on vectors:

```{r}
mean(d) # calculate the mean of the vector
sd(d) # the standard deviation
```




