---
title: "Foundations of R programming I"
author: "Shamit Soneji"
date: "12/10/2018"
output: rmarkdown::github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Why use R?

R is a programming environment with a focus on mathematics and statistics, but can be used for a variety of applications given the flexibility of the language. R is also free and available for all operating systems. Given the richness of the language and no cost to use it, bioinformaticians have adopted R as the platform for which which to develop packages to solve bioinformatics problems.

## Getting R

Point your browser to http://cran.r-project.org/ to download and install the latest version of R. For these tutorials we are also going o use [RStudio](http://www.rstudio.com/) which is an advanced environment for R which includes a window for an editor, console, and plotting window. You will see what this means later.

With respect to bioinformatics, the central repository for bioinformatics tools is [Bioconductor](http://www.bioconductor.org) where packages are deposited for easy installation.

Before we go on to these, we need to get to grips with the basics of the R language first.

### Before we do anything, go to Session > Set Working Directory > Choose Directory and find the "RIworkshop" I asked you to make in an earlier email.

## R- The basics

### 1- Vectors

The R language is relatively intuative. For example, making a string of numbers from 1 to 10:

```{r}
x <- c(1,2,3,4,5,6,7,8,9,10)
```
The "c" in the code above means "combine", therefore all the comma separated numbers between the parentheses are put together to create a `vector`.

To view the contents of the object you have just created, just type "x" and hit return:

```{r}
x
```

There is a much simpler way to create the same type of object:

```{r}
x <- 1:10
x
```
Much better. Using a comma will always do increments of 1, but is also bidirectional:

```{r}
y <- 5:-5
y
```

Another way of ceating a sequence of numbers is to use the `seq` function. To learn how this function works, issue the command `help(seq)`. In R you can get a manual for any function using the `help()` command. To generate a vector of numbers from 1 to 100 in steps of 10 we need:

```{r}
a <- seq(0,100,by=10)
a
```
***Exercise:*** Generate a vector called 'b' ranging from 3 to 987 where the length of the vector is 53 entries long.
Done? Check the length of the vector you have just made by issuing `length(b)`.

Now that we can make vectors we can start playing with them. for example:
```{r}
c <- 1:50
d <- 1/c
```
Lets plot the numbers contained in the object we called `d`:

```{r fig.width=7, fig.height=6}
plot(d)
```
Note the way the axes are labelled in the plot function.

***Exercise:*** Call `help(plot)` and read about the other options available. Produce the same plot as above, but this time as a line plot which is coloured red. Also, label the axes and give the plot a title.

We can also do basic calculations on vectors:

```{r}
mean(d) # calculate the mean of the vector
sd(d) # the standard deviation
```

### 2- Matricies

Matricies are the most common data format bioinformaticians work with (microarray data for example). Lets make one:

```{r}
m <- matrix(0,ncol=5,nrow=10)
m
```
This will create a matrix filled with zeros. To transpose the matrix we use `t()` (this will be important later)

```{r}
tposed.m <- t(m)
tposed.m
```
Lets make a matrix (and a vector) containing numbers so we can look at subsetting:
```{r}
v <- 1:10
m <- t(matrix(1:50,ncol=10,nrow=5))
m
```
We can access individual elements using square brackets `[]`. Here are some examples:

```{r}
v[c(7,1,5)] #access elements 7 1 and 5 of the vector
m[1,] # access the first row of the matrix
m[,3] # the 3rd column
m[8,2] # the value in the 8th row and 2nd column
m[3:7,4] # the 3rd to 7th row of the 4th columnß
```
We often have the need to collect vectors and assemble them into matricies. This can be done using the `rbind` (row) and `cbind` (column) function:

```{r}
v1 <- 1:10
v2 <- 101:110
rbound.mat <-rbind(v1,v2)
cbound.mat <- cbind(v1,v2)
rbound.mat
cbound.mat
```

### 3- Lists

Lists are a class of container which allow multiple objects to be stored in a single variable. For example:
```{r}
alpha <- LETTERS[1:8]
mat <- matrix(rnorm(40),nrow=8)
listex1  <- list(char=alpha,nums=mat)
```
Each element can now be accessed via `$`:
```{r}
listex1$char
listex1$nums
listex1$nums[1,] # the matrix within the list issubsetted as before
```
Another way of doing the above is:
```{r}
listex1[[1]] # note the double square brackets
listex1[[2]]
listex1[[2]][1,]
```

4- Reading and writing files
==========================
To analyse data in R, you have to get the data into R first (this helps a lot). R has some useful functions to do this, so now we can take our first look at some microarray data. Download this file (http://bone.bmc.lu.se/Public/Mouse_HSPC_reduced.txt) and save it to your current working directory.

***Exercise:*** Open the file to see how it looks, and then call `help(read.delim)` and try to work out how the file you are looking at could be read into R.

This is how I would do it:

```{r}
hspc.data <- read.delim("Mouse_HSPC_reduced.txt",header=T,row.name=1,sep="\t")
colnames(hspc.data) # view the column names
nrow(hspc.data) # the number of rows in the dataset
ncol(hspc.data) # number of columns
dim(hspc.data) # number of rows and columns together
colnames(hspc.data) #output the columns headers
```

***Exercise:*** Using subsetting we learnt about earlier, split this data matrix into three parts called `lthsc`, `mep` and `gmp` to separate the cell types shown in the headings. For this look at the help page for a function called `grep`.

```{r}
lthsc <- hspc.data[,grep("LTHSC",colnames(hspc.data))]
mep <- hspc.data[,grep("MEP",colnames(hspc.data))]
gmp <- hspc.data[,grep("GMP",colnames(hspc.data))]
```

To write a table use the `write.table` function:
```{r}
write.table(lthsc,"LTHSC_data.txt",row.names=T,col.names=NA,sep="\t",quote=F)
```

***Exercise:*** Write out the data for the MEP and GMP data into two files.

The data tables we have now are in the form of a data.frame. Try:

```{r}
class(mep)
```
This can be an awkward format for some operations so we can conver it to a simple matrix first:

```{r}
hspc.data <- as.matrix(hspc.data)
lthsc <- as.matrix(lthsc)
mep <- as.matrix(mep)
gmp <- as.matrix(mep)

```

Try this now:
```{r}
class(mep)
```

### 5- Flow control and Functions

This is where it starts to get more interesting. Now we have these two datasets loaded in memory, we can use them to learn about flow control and some basic mathematical functions. We are going to do a few things the "long way" so you get the idea of how flow control works, and then we'll look at some shortcuts.

Flow control is how multi-step processes are carried out. In the example below we print out the numbers 1 to 10:

```{r}
for(i in 1:10){
  print(i)
}
```
To translate this code, it simply says for every integer from 1 to 10, print this value to the screen.

***Exercises:*** 
- Using the example above, print the first 10 lines of `lthcs` in a `for` loop.
- Print every 2nd line of `mep` from lines 1 to 50.

An important point regarding `for` loops is that any processes/calculations occurring within the loop will stay in the loop. If data generated within a loop has to be retained, we need to create a container to "fill up" while the loop is being carried out.

```{r}
vec <- NULL
for(i in 1:10){
  vec <- c(vec,i*10)
}
```
The container `vec` is initialised outside the loop, and then populated by concatenating on to it after every iteration of the loop.

***Exercise:*** Initialise an empty container, and for `gmp`, calculate the mean of each row (gene), and store the results in the containers you made.

```{r}

gmp.row.mean <- NULL

for(i in 1:nrow(gmp)){
  gmp.row.mean <- c(gmp.row.mean,mean(gmp[i,]))
}
```

We will now take a look at *functions*. As an example, lets turn the code you wrote in the exercise above into a function where we also calculate the variance for a gene too.

```{r}

calc.mean.and.sd <- function(mat){
  
  mn <- NULL
  vr <- NULL
  
  for(i in 1:nrow(mat)){  
      mn <- c(mn,mean(mat[i,]))
      vr <- c(vr,var(mat[i,]))
    
  }
  res <- list(mns=mn,vars=vr)
  res # the last line in a function is what the function will return
}

```
By putting this code into a function we can now calculate the means and deviations of any matrix. For example, `gmp`:

```{r}
gmp.mn.sd <- calc.mean.and.sd(gmp) 
```

Functions can also work with built in conditions:

```{r}
animal.maths <- function(value1,value2,animal=c("pig","cow")){
  
  if(animal=="pig"){print(value1/value2)}
  if(animal=="cow"){print(value1*value2)}
  
}
animal.maths(5,5,"pig")
animal.maths(5,5,"cow")

```



These functions can now be "banked" for use whenever they are needed (probably not `animal.maths` to be fair). However, you should avoid using for-loops etc altogether since R has some built in functions that are much quicker and tidier. Lets look at that now.

### 6- Apply

'apply' is a commonly used function in R to speed up matrix calculation. For example, to calculates means of a matrix we can do this:

```{r}
lthsc.row.mn <- apply(lthsc,1,mean) # means of rows
lthsc.col.mn <- apply(lthsc,2,mean) # means of columns
```
The format for the function is therefore the matrix first, the direction in which you would like to apply the function in the 3rd argument.

 use `apply` to calculate row and column totals and deviations for a yeast dataset of your choosing.

Your own functions can also be used with `apply` when used as the 3rd argument. Example:

```{r}

example.func <- function(v){
  
  val <- (mean(v)*sd(v))/sum(v) ## This is a nonsense operation.
  val
}

ex.apply <- apply(mep,1,example.func)

```

Lets use the apply function to get the top 500 most variable genes in our HSPC dataset:
```{r}
gene.vars <- apply(hspc.data,1,var)
top.var.genes <- names(rev(sort(gene.vars))[1:500])
hspc.var <- hspc.data[top.var.genes,]
```
### 7- Standardising data

Lets take a look at the range of the data, i.e getting the lowest and higest value in the matrix of variable genes we just made.
```{r}
range(hspc.var)
```


For some operations (such as making heatmaps) the data needs to be z-score normalised (scaled) first. When we scale data, each row of gene is standarised so that it's mean=0 and sd=1. Specifically for a gene `g` of the i-th row:

$$Z_i= \frac{g_i-\hat{g}}{\sigma_g}$$

which means for each row we subtract the mean of the row, and then divide by the sd of the row.

***Exercise:*** write a function called `zscore` which will take a single vector of values and scale them. When you have done this, `apply` this to the `hspc.var` matrix to scale all rows and call it `hspc.zs`.
```{r, echo=FALSE}
zscore <- function(v){
  z <- (v-mean(v))/sd(v)
  z
}

hspc.zs <- apply(hspc.var,1,zscore)
```

Now take a look at the first row of the normalised data. Call `nrow` on the matrix. Does it look right? 

```{r}
hspc.zs <- t(apply(hspc.var,1,zscore))
boxplot(hspc.zs,las=2)
# compare to the original data
boxplot(hspc.var,las=2)
```

We can see now the data has been centralised around 0.

8- Clustering
====================

Clustering is one of the most common visualisation techniques for genes expression data. Here we will learn how to do some basic histograms/heatmaps and plotting. The first thing  
R has many ways to do this, and many packages have been written specifically for expression data. We are not going to use these for now, but concentrate on the basic underlying functions that do the maths. For example, the `gplots` package uses the `hclust` function which is provided by R. So we will use `hclust` for now.

To use `hclust` we need to provide a distance matrix. This is done using the `dist` function:

```{r}
hspc.dst <- dist(hspc.zs)
```

The we cluster using `hclust`:

```{r}
hspc.hc <- hclust(hspc.dst)
```

Plot the dendrogram:

```{r}
plot(hspc.hc)
```

You'll see form this what we have clustered are the genes. If you want to cluster the cells then you need to transpose the matrix using `t()`:
```{r}
hspc.dst <- dist(t(hspc.zs)) #transpose the matrix here!
hspc.hc <- hclust(hspc.dst)
plot(hspc.hc)
```

We can see this is pretty much useless. It is far to compact and doesn't really tell us anything. What we would like is to make a heatmap where the genes and samples are clusters, and to do this we need to retrieve some information created by `hclust`

Call `names` to see which information is available in the newly created object:

```{r}
names(hspc.hc)
```

What we need here is the component called `order`. We can get this using the `$` assignment.

```{r}
hspc.hc$order
```

This is the order the cells appear in form left to right when you plotted the dendrogram os cells ust before. We use this to reorder the z-scored matrix:

```{r}
hspc.cell.clustered <- hspc.zs[,hspc.hc$order]
```

To make a heatmap of the data call `image`:

```{r}
image(hspc.cell.clustered)
```
Ok, this doesn't look like it should! The matrix is the wrong way round, the colours aren't right, and there are no labels. The downside to R is that getting all this done takes time and knowledge of R's plotting capabilities. Thankfully people have already done this and put the code into functions/packages for people to download and use.

***Exercise:*** Install the `pheatmap` package.

To use the functions provided by gplots we have to load it first:

```{r}
library(pheatmap)
```

We can now use the `pheatmap` function that the package provides:
```{r}
pheatmap(hspc.zs)
```

The pheatmap function uses `hclust` to cluster the genes and cells and reorders the matrix according to both. Lets output this to a file:

```{r}
png("HSPC_heatmap.png",height=4500,width=1500)
pheatmap(hspc.zs)
dev.off()
```
The file is opened, and the plot is then made. The `dev.off()` then closed and finalises the file, i.e nothing more can be written to it.

***Exercise:*** call `help (pheatmap)` and see what options are available. Play with the options to see what they do.

Clustering is pretty pointless if you can't define groups and get to the gene names. First we need to capture the output from `pheatmap` as a variable:

```{r}
hspc.clust <- pheatmap(hspc.zs)
```
Lets take a look at the contents of `hspc.clust`:
```{r}
names(hspc.clust)
```

What we want is the information contained within the hclust object in `tree_row`. We get this by treating it like a list:

```{r}
hspc.clust$tree_row
hspc.clust$tree_row$order #the order of genes in the heatmap for examples.
```

Lets say that we want to split the genes in to 5 clusters groups, we can call the `cutree` function on an `hclust` object to do this:
```{r}
gene.clusters <- cutree(hspc.clust$tree_row,k=5)
gene.clusters[1:20] # shows the results for the first 20 genes.
table(gene.clusters)
barplot(table(gene.clusters))
```

Lets isolate all the genes beloning to cluster 1 using the `which` command:

```{r}
which(gene.clusters==1)
```
We can isolate these rows only from our `hspc.zs` matrix as we did before:

```{r}
hspc.cluster.1 <- hspc.zs[names(which(gene.clusters==1)),]
```

We can now see how these gene behave as a whole using a boxplot:

```{r}
boxplot(hspc.cluster.1,las=2)
```
